Synthesizing Quotient Lenses

### 2018-03-10 papers/QuotientLenses/main.tex
Merged Dave's work back in

\begin{document}
\title{Synthesizing Quotient Lenses}
\begin{abstract}
There currently exist many bidirectional programming languages, some of which
allow a programmer to define a transformation that respects an equivalence
relation. For example, a language with this feature may allow a programmer to
define a transformation which ignores the order of fields in an XML file, so
that two files that have the same fields with the same data but with the fields
occuring in a different order are considered the same by the
transformation. This paper simplifies the task of programming such {\em
quotient lenses} in two main ways. First, we introduce {\em
quotient regular expressions} or QREs which are regular expressions
augmented with extra syntax that makes it easier to specify equivalence
relations on regular langauges than currently existing tools. Secondly, we
introduce {\em QRE lenses}, which are lenses that map between QREs, and show
how to {\em synthesize} QRE lenses from a pair of QREs and a set of example
input-output pairs. We also describe an implementation of QREs and QRE lens
synthesis in the bidirectional programming language Boomerang, and demonstrate
the utility of our implementation by using it to synthesize QRE lenses between
various real-world data formats used by the {\tt data.gov} database.
\end{abstract}

@@

\section{Introduction}
Oftentimes, the same data is stored in a variety different formats. For
example, bibliography data may be stored in \bibtex{} and
EndNote formats; spreadsheet data can be represented as a CSV file
or a TSV file; web APIs can return data as a JSON or XML object. In such cases,
it is necessary to synchronize the two or more copies of data so that they
store the same information.

One way to synchronize data stored in different formats is to use \emph{lenses},
\emph{i.e.,} programs that represent pairs of functions for translating data
from format the source format to the target formant (the \emph{get}
direction) and back again (the \emph{put} direction). Lenses are usually
derived using domain-specific bidirectional languages which guarantee that the
lenses defined in that language satisfy a certain set of \emph{lens laws} that
give guarantees on how the two transformations interact.

Amongst the most restrictive lenses are the \emph{bijective lenses}.
A bijective lens $\ell : S \Leftrightarrow T$ is simply a bijection from
a source set $S$ to a target set $T$:
\begin{equation}\label{bijectivelenslaws}
\ell.\get \; (\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s)
= s
\end{equation}
Since these lenses demand that get and put functions form a bijection, then
no information is lost when one translates from one data format to another. 
While this strong guarantee can be useful in some contexts, it is more often
the case that some of the information in one or both of the data formats is
superfluous. For instance, one typically does not care to preserve the amount of
whitespace between characters.  In other circumstances, capitalization
or the order of fields in a record or items in a list may be unimportant. Many
data formats have other ad hoc rules as well. In bibliographies, for example,
recall that in \bibtex{}, an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---it does not
matter which of the two formats is used.

{\em Quotient lenses}~\cite{quotientlenses} loosen many of the restrictions of
bijective string lenses by allowing a programmer to define an equivalence
relation on the source and target formats so that the round-tripping lens laws
hold modulo these equivalence rather than modulo equality. Quotient lenses thus
make it possible to align many more data sources than would otherwise be
possible, by ignoring details that a programmer views as inessential. The only
language we are aware of that implements a general class of quotient lenses is
{\em Boomerang}~\cite{boomerang}. While Boomerang enables a programmer to
define a very general class of quotient lenses, these lenses to have two main
drawbacks.

First, in order to define an equivalence relation in Boomerang, a programmer
must define a {\em canonizer}; pair of functions $f$ and $g$ where $f$ maps
data represented in the original external format to a canonical element and $g$
chooses which elements of the original format correspond to values in the
canonical format. The disadvantage of this approach is that in order for a
programmer to {\em specify} a quotient lens in Boomerang, she must define four
different data formats: two ``outer'' formats which specify the data that is to
be transformed by the lens, and two ``inner'' formats which specify the
canonical formats for the source and target data.

Our solution to this shortcoming is to introduce {\em Quotient Regular
Expressions} or QREs, which enable a programmer to simultaneously specify a
regular expression as well as a set of canonical representatives for strings
that match the regular expression. QREs have the added advantage that once the
programmer has defined a well formed QRE, then she need not define a
canonization function seperately; the canonization function can automatically be
inferred from the QRE.

The second challenge with programming lenses in Boomerang is that even after the
programmer has specified the data as well as a canonizer that maps each member
in the source and target sets to their respective canonical representative, she
must still define the underlying lens that maps between the respective sets of
canonical representatives. This can be a slow and tedious task due to the
constraints that Boomerang imposes on its terms. For example, Boomerang
programmers often must rearrange the order of data items by recursively using
operators that \swap adjacent fields. Furthermore, the Boomerang type checker is
very strict, disallowing many programs because they contain ambiguity about how
certain data is transformed.

Our solution to this problem is that we use QREs to define a class of quotient
lenses called {\em QRE lenses}, and, we show how to synthesize such QRE lenses
automatically. More specifically, in order to generate a lens, a programmer
provides a pair of QREs describing the source and target formats, plus
a (possibly empty) suite of examples that demonstrates how the
lens should act in specific cases.  Given this information, we can exploit and
extend recent work on algorithms for synthesis of bijective
lenses~\cite{optician} to generate quotient lenses. The lenses generated by
our synthesis algorithm are guaranteed to translate data back and forth
between source and target QREs and to satisfy the provided examples.
Moreover, while our synthesized quotient lenses have a relatively
restricted form (which cuts down the amount of non-determinism in
a specification and greatly speeds up synthesis), we prove that
this rigidity does not compromise their expressiveness.

The following list summarizes the central contributions of our work.
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs),
  a compact, convenient notation for simultaneously defining a
  regular language modulo some equivalence relation and a canonizer
  for that relation (Section~\ref{QRE}).
  \item We introduce {\em QRE lenses} which are lenses that map data specified
  using QREs.
  In addition, we will descibe a particular set of QRE lens combinators and
  prove that QRE lenses defined using these combinators can be transformed to
  a QRE lens that us defined using just two of these combinators; this is the
  main technical contribution of this paper.
  (Section~\ref{QRE-lenses}).
  \item We use this result to reduce the problem of {\em synthesizing} QRE
  lenses to the problem of synthesizing bijective string lenses which we studied
  in~\cite{optician}(Section~\ref{synth}).
  \item We extend the Boomerang {\em implementation} with QREs
  and QRE lens synthesis.  We demonstrate the utility of our implementation by
  using it to synthesize QRE lenses between data formats drawn from the
  {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

\bcp{We've said that one of the main contributions of the paper is
synthesizing lenses, but we don't actually show this happening in the
introduction.  Should we?}

\section{QRE Lenses by Example}
\label{sec:example}
\subsection{Bijective String Lenses}
Consider a bidirectional transformation that converts \bibtex{} citation records
such as
\begin{verbatim}
@Book {Doe,
Author = "John Doe",
Title = {Generic Title}
}
\end{verbatim}
\noindent
into an equivalent EndNote format like the following
\begin{verbatim}
%0 Book
%F Doe
%A John Doe
%T Generic Title
\end{verbatim}
\noindent
and vice versa.

@@

\subsection{Quotient Lenses}
Unfortunately, while our synthesis tool greatly simplifies the task of
programming bijective string lenses, not all bidirectional tranformations are bijective
in nature. For instance, users are typically uninterested in preserving
whitespace between words or between fields.  The order of author and title
fields is also likely irrelevant and there may be equivalent ways of writing
the same name:  ``Doe, John'' vs ``John Doe.''

When two instances of the same format differ only terms of such inessential
information, it is useful\bcp{Would it be worth spelling out the alternative
(i.e., what lenses would look like if they had to preserve all this stuff?)}
to treat them as identical and to map them into canonical representations.
Continuing our example, the following two \bibtex{} citations represent the same
logical object and as a result, should be translated to the same EndNote record.
\begin{verbatim}
@Book {Doe,
Author = "John Doe",
Title = {Generic Title},
}

@Book{     Doe,
Title = {Generic
Title},
Author = "Doe, John",     }
\end{verbatim}

@@

Sometimes there are multiple disjoint representations of the same data.
In such situations, the \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in the first format to the second.
For instance, assume that
the \codefont{comma\_name} format describes ``Doe, John''
and that the \codefont{space\_name} format describes ``John Doe''
and \codefont{c\_to\_s} is a function from the first to the second.  In this
case, the following instance of \kw{squash} creates the desired canonizer.

@@

One way to define the \codefont{c\_to\_s} canonization function used by squash
is simply to write it from scratch in some ordinary programming language.
However, our system will also synthesize such functions automatically from pairs
of regular languages and examples---here, \codefont{c\_to\_s} is the
\codefont{get} direction of a lens that can be synthesized by:
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Doe, John", "John Doe")}
let c_to_s = get l
\end{lstlisting}

@@

(* synthesize a lens that maps comma representation to space representation *)
let l = synth comma_name $\Leftrightarrow$ space_name using {("Doe, John", "John Doe")}
let c_to_s = get l

@@

Now let $Q_1$ and $Q_2$ be QREs. Given a bijection $\ell : K(Q_1)
\Leftrightarrow K(Q_2)$, we may define a quotient lens $q : W(Q_1)/\eqrel{Q_1}
\Leftrightarrow W(Q_2)/\eqrel{Q_2}$ by $q.\get = \ell \circ \canonizer(Q_1)
$ and $q.\lput = \ell_1^{-1} \circ \canonizer(Q_2)$. In other words, the get
function of $q$ first canonizes the source data using $\canonizer(Q_1)$
and then applies $f$ to the result, while the put function of $q$ first
canonizes the target data using $\canonizer(Q_2)$ and the applies $f^{-1}$ to
the result. This is our approach in defining QRE lenses.

One potential pitfall with QRE lenses is that a lens derived using the QRE lens
combinators may not have the form of a bijection with ``canonizers at the
edges''. Such a situation would be undesirable since our strategy for
synthesizing QRE lenses is to synthesize a bijective string lens which plugs
into the two QREs provided as input to the synthesis procedure; if there were
some QREs which could not be transformed to the desired form, then our synthesis
algorithm would necessarily be incomplete.

Thankfully, this turns out not to be an issue if we use the bijections that
can be expressed in Boomerang; following ~\cite{optician} we call these
lenses {\em bijective string lenses}. Once we restrict the bijections that we
use to derive QRE lenses to include just bijective string lenses, we will be
able show that every QRE lens can be transformed to a QRE lens which has the
form of a bijective string lens with canonizers at the edges.

\subsection{Bijective String Lenses}
The syntax for the language of bijective string lenses is given by $$\ell_1 :=
\mathit{id} \; (R) \sep \const(s_1, s_2) \sep  \swap(\ell_1, \ell_2) \sep
\ell_1 \cdot \ell_2 \; |  \; (\ell_1 \sep \ell_2) \sep \ell^* \; | \; \ell_1 \;
; \;  \ell_2,$$ where $R$ ranges over regular expressions and $s$ ranges over
character strings.

The denotation of a lens $\ell_1$ is $\llbracket \ell_1 \rrbracket \subseteq
\mathit{String} \times \mathit{String}$. If $(s_1, s_2) \in \llbracket \ell_1
\rrbracket$, then $\ell_1$ maps between $s_1$ and $s_2$.
\begin{align*}
\llbracket R \rrbracket &= \{(r, r) \sep r \in \mathcal{L}(R)\}\\
\llbracket \const(r, s) \rrbracket &= \{(r, s)\}\\
\llbracket \swap(\ell_1, \ell_2) \rrbracket &= \{(s \cdot t, t' \cdot s') \sep
(s, s') \in \llbracket \ell_1 \rrbracket \text{ and } (t, t') \in \llbracket
\ell_2 \rrbracket\}\\
\llbracket \ell_1 \cdot \ell_2 \rrbracket &= \{(s \cdot t, s' \cdot t) \sep
(s, s') \in \llbracket \ell_1 \rrbracket \text{ and } (t, t') \in \llbracket
\ell_2 \rrbracket\}\\
\llbracket \ell_1 \sep \ell_2 \rrbracket &= \{(s \cdot t) \sep
(s, t) \in \llbracket \ell_1 \rrbracket \text{ or } (s, t) \in \llbracket
\ell_2 \rrbracket\}\\
\llbracket \ell^* \rrbracket &= \{(s_1 \cdot \ldots \cdot s_n, t_1 \cdot \ldots
\cdot t_n) \sep (s_i, t_i) \in \llbracket \ell \rrbracket \text{ for } 1
\leq i \leq n\}\\
\llbracket \ell_1 \; ; \; \ell_2 \rrbracket &= \{(r, t) \sep \text{there exists }s
\text{ such that } (r, s) \in \llbracket \ell_1 \rrbracket \text{ and } (s, t)
\in \llbracket \ell_2 \rrbracket\}
\end{align*}
The $\mathit{\const}(s, t)$ lens replaces the string $s$ with $t$ in the source
data in the forward direction, and $t$ with $s$ in the target data in the backward
direction, while $\mathit{id}(R)$ applies the identity function to the
source and target  in $\mathcal{L}(R)$ in both directions. The composition lens
$\ell_1 \; ; \; \ell_2$ applies $\ell_1$ followed by then $\ell_2$ to the
source in the forward direction, and applies $\ell_2$ followed by $\ell_1$ to
the target in the backward direction. The lens $\ell_1 \cdot \ell_2$ first
splits the string $s$ into $s_1$ and $s_2$, applies $\ell_1$ and $\ell_2$ to
$s_1$ and $s_2$ to get $t_1$ and $t_2$ respectively, then concatenates $t_1$
and $t_2$ and returns $t_1 \cdot t_2$ as the final result in the forward
direction. $\ell_1 \cdot \ell_2$ operates similarly in the backward direction,
but with $s, s_1$ and $s_2$ substituted for $t, t_1$ and $t_2$. The
$\mathit{\swap} \; (\ell_1, \ell_2)$ lens operates like $\ell_1 \cdot \ell_2$,
except that it swaps $t_1$ and $t_2$ before concatenating the two for a final
result of $t_2 \cdot t_1$ in the forward direction. In the backward direction,
$\mathit{\swap}(\ell_1, \ell_2)$ first undoes the \swap, then behaves like
$cdot$. The $\ell_1 \; | \; \ell_2$ lens chooses to apply $\ell_1$ or $\ell_2$
depending on whether the source (resp. target) data is matched by $\ell_1$ or
$\ell_2$ in the forward (resp. backward direction). The $\ell^*$ lens splits the
string $s$ into strings $s_1, \ldots, s_n$, applies $\ell_1$ to each $s_i$ to
get $t_i$, and then concatenates each of the $t_i$'s for a final result of $t_1
\cdot \ldots \cdot t_n$ in the forward direction. The iteration lens $\ell^*$
operates similarly in the backward direction, but with $s, s_i$ substituted for
$t, t_i$.

Each bijective string lens $\ell$ has a type $\ell : R \Leftrightarrow S$ where
$R$ and $S$ are regular expressions. If $\ell : R \Leftrightarrow S$, then the
source language of $\ell$ is $\mathcal{L}(R)$ and the target language of $\ell$
is $\mathcal{L}(S)$. Interestingly, with the bijective string lens type system,
a lens $\ell : R \Leftrightarrow S$ can also be considered to be of type $\ell
: R' \Leftrightarrow S'$ provided that $R$ (resp. $S$) can be proven to be
equivalent to $R'$ (resp. $S$) from the star-semiring axioms (associativity and
commutativity of | and $\cdot$, identity of the empty string $\epsilon$,
distibutivity of $\cdot$ over |, annihilative property of $\varnothing$ with
respect to $\cdot$, and that $R^* = \epsilon \; | \; R \cdot R* = \epsilon \; |
\; R^* \cdot R$ for all $R$):

@@

The other typing rules for bijective string lenses are given in
Figure~\ref{fig:lensrules}.

@@

\subsection{Syntax of QRE Lenses}
Having given a brief overview of the class of bijective lenses, we now introduce
the class of QRE lenses. The language of QRE lenses is given by following
grammar:
$$ q := \kw {id}(R) \sep  \kw{const}(r, s) \sep q_1 \cdot q_2 \sep
\kw{swap}(q_1, q_2) \sep (q_1 \sep q_2) \sep q^* \sep q_1 \; ; \; q_2 \sep
\kw{lquot}(q, \ell) \sep \kw{rquot}(\ell, q)$$ where $R$ ranges over
regular expressions and $\ell$ ranges over bijective lenses.

@@

\iffalse
The difference between QRE lenses and Boomerang quotient lenses lies in the
typing judgements. More concretely, each QRE lens $q_1$ has a type $q_1 : c
\Leftrightarrow Q_1$ where $c, Q_1$ are QREs. In contrast, the approach used to
type quotient lenses in Boomerang is to classify equivalences according to
whether they are or are not the equality relation. This type system is based on
two observations: first, that most quotient lenses originate as lifted basic
lenses, and therefore have types whose equivalence relations are both equality;
and second, that equality is preserved by many of the quotient lens
combinators. Foster et al also discuss a second possible approach to typing
quotient lenses, where equivalence relations are represented by rational
functions that induce them. While this second approach is more refined than the
first, Boomerang favours the first approach since the second appears to be too
expensive to be useful in practice~\cite{quotientlenses}.
\fi

\subsection{Semantics of QRE Lenses}

Each QRE lens $q$ has a type $q : Q_1 \Leftrightarrow Q_2$ where $Q_1, Q_2$ are
QREs. If $q : Q_1 \Leftrightarrow Q_2$, then the source format is described by
$W(Q_1)$ and the canonical set of representative for the source data is
described by $K(Q_1)$.
Similarly, the target format is described by $W(Q_2)$ and the canonical set of
representatives for the target data is described by $K(Q_2)$. The underlying
lens of $q$ is a bijective lens $\ell : K(Q_1) \Leftrightarrow K(Q_2)$.

>>>>

\begin{document}
\title{Synthesizing Quotient Lenses}
\begin{abstract}
{\em Quotient lenses} are bidirectional transformations that respect
specified equivalence relations with the goal of 
appropriately handling inessential details in data formats. 
For example, a programmer could use a quotient lens to define 
a transformation that ignores the order of fields in XML data, so
that two XML files with the same fields but in different orders would be
considered the same. 
This paper simplifies the task of programming quotient lenses in three
steps.
First, we introduce {\em quotient regular expressions} (QREs) to
make it easy to express which portions of a data format are
inessential.  A QRE is an augmented regular expression that
simultaneously defines a regular expression and an equivalence
relation over matching strings.  Variations within the same
equivalence class are deemed inessential. 
Next, we introduce {\em QRE lenses}, which are lenses that map
between QREs.
Our key technical result is a proof that every QRE lens can be
transformed into a QRE lens that has the form $c_1 \cdot \ell \cdot
c_2$, where $c_1$ and $c_2$ are {\em canonizers} 
and $\ell$ is a bijective transformation.  Intuitively,
the canonizers normalize the data  by picking a
canonical representative of the relevant equivalence relations. 
This theorem means we can push all the normalization to the edges of a
bidirectional transformation, leaving in the middle a pure bijective
transformation.  
Finally,
we levarage this insight to {\em synthesize} QRE lenses from a pair
of QREs and example input-output pairs by reusing earlier work on
synthesizing bijective lenses.
We have implemented QREs and QRE lens synthesis 
in the bidirectional programming language Boomerang.
We evaluate the effectiveness of our approach by using it to
synthesize QRE lenses between various real-world data formats in
the {\tt data.gov} database.

\end{abstract}

@@

\section{Introduction}
Often, the same data is stored in more than one format. For example,
bibliographic data may be stored in \bibtex{} or EndNote formats,
spreadsheet data can be represented as either CSV and TSV files, and web
APIs can return data as either JSON or XML objects.  When changes are
made to data in one format, data in the other must be updated so the
two representations continue to store the same information.
One way to synchronize such data is to use \emph{lenses}\ksf{include citations?}, which
are programs that simultaneously define pairs of functions for translating
the data 
from the source format to the target formant (the \emph{get}
direction) and back again (the \emph{put} direction). Lenses are usually
defined using domain-specific bidirectional languages that 
ensure the specified lenses satisfy various \emph{lens laws} that
guarantee how the \emph{put} and \emph{get} functions interact.

\emph{Bijective lenses} are amongst the most restrictive. 
A bijective lens $\ell : S \Leftrightarrow T$ specifies a bijection from
a source set $S$ to a target set $T$:
\begin{equation}\label{bijectivelenslaws}
\ell.\get \; (\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s)
= s
\end{equation}
Since these lenses demand that the \emph{get} and \emph{put} functions form a bijection, 
no information is lost during translation from one data format to another. 
While this very strong guarantee can be useful, more often
some of the information in one or both of the data formats is
superfluous. For instance, the number of white space characters
between data items, the capitalization of various strings, the
sequence of fields in record, or the order of items in a list are
often irrelevant.  Many
data formats have their own ad hoc rules as well. In \bibtex{}, for example,
an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---
the specific choice does not matter.

{\em Quotient lenses}~\cite{quotientlenses} loosen the restrictions of
bijective string lenses by including equivalence
relations on the source and target formats so that the round-tripping lens laws
hold modulo these equivalence rather than modulo equality. Quotient lenses thus
make it possible to synchronize many more data formats than would otherwise be
possible by ignoring details deemed inessential.

The bidirectional programming language {\em
Boomerang}~\cite{boomerang} is the only language we know of that
implements a general class of quotient lenses.  Unfortunately, these
lenses to have two main drawbacks.

The first drawback is that defining an equivalence
relation in Boomerang requires specifying a {\em canonizer}, which is
a pair of functions $f$ and $g$ where $f$
normalizes data from the
external format by choosing canonical representations
and $g$ provides an
inverse by specifying the external form of each canonical
representation. 
The disadvantage of this approach is that writing
a quotient lens in Boomerang requires defining four
different data formats: two ``external'' formats which specify the data to
be transformed by the lens, and two ``internal'' formats which specify the
canonical representations for the source and target data.

Our solution to this shortcoming is to introduce {\em Quotient Regular
Expressions} (QREs), which enable programmers to simultaneously specify a
regular expression and a set of canonical representatives for matching
strings.  Given an QRE, we can automatically infer the corresponding
canonization functions.  
\ksf{Define comma\_name here as a really simple example? }

The second drawback with writing quotient lenses in Boomerang is that
even after specifying the four required data formats, the programmer must
still define a lens for converting between the canonical
representations.   Experience shows that writing such lenses can be
tedious 
(consider having to rearrange the order of data items by recursively
using operators to \swap adjacent fields) 
and that satisfying the type checker can be difficult because 
many seemingly useful lenses contain ambiguities disallowed by the 
Boomerang type checker~\cite{optician}.

Our solution to this problem is to introduce {\em QRE lenses}, which
are a new class of quotient lens that can be automatically
synthesized, obviating the need to write lenses by hand.  A QRE lens
is like a normal lens except it maps between formats described by
QREs. We prove a normal form theorem (Theorem~\ref{normal form}) that says that every
composition of QRE lenses can be rewritten 
to be the composition of a source canonizer, a bijective lens, and a
target canonizer. This normalization property means we can synthesize
QRE lenses by extending an existing algorithm for synthesizing
bijective lenses~\cite{optician}.

Given this framework, generating a QRE lens requires only a pair of QREs to describe the
source and target formats and a (possibly empty) suite of examples
demonstrating the mapping.  For example, the following code

\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Lovelace, Ada", "Ada Lovelace")}
\end{lstlisting}
binds \codefont{l} to a synthesized QRE lens mapping between names in
the comma-separated form describd by the QRE \codefont{comma\_name}
and the space-separated form described by the
QRE \codefont{space\_name}. 
The lenses generated by
our synthesis algorithm are guaranteed to translate data back and forth
between source and target QREs and to satisfy the provided examples.
Our synthesized quotient lenses have a relatively
restricted form (which greatly speeds up synthesis), but the 
normalization theorem means 
this restriction does not compromise expressiveness.

The following list summarizes the contributions of our work.
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs), which are
  a compact, convenient notation for simultaneously defining a
  regular language modulo some equivalence relation and a canonizer
  for that relation (Section~\ref{QRE}).
  \item We introduce {\em QRE lenses}, which are lenses that translate
  between data formats specified using QREs. (Section~\ref{QRE-lenses}).
  \item We descibe a particular set of QRE lens combinators and prove
  that QRE lenses defined using these combinators can be normalized
  (Theorem~\ref{normal form}), which is the main technical
  contribution of this paper.
  (Section~\ref{QRE-lenses}).
  \item We use this result to reduce the problem of {\em synthesizing} QRE
  lenses to the problem of synthesizing bijective string lenses
  previously studied in~\cite{optician}(Section~\ref{synth}).
  \item We extend Boomerang with QREs and QRE lens synthesis. 
  We use the resulting implementation to demonstrate the utility of
  our approach by synthesizing QRE lenses between
  data formats drawn from the {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

\section{QRE Lenses by Example}
\label{sec:example}
\subsection{Bijective String Lenses}
Consider a bidirectional transformation that converts \bibtex{} citation records
such as
\begin{verbatim}
@Book {Lovelace,
Author = "Ada Lovelace",
Title = {Generic Title}
}
\end{verbatim}
\noindent
into an equivalent EndNote format like the following
\begin{verbatim}
%0 Book
%F Lovelace
%A Ada Lovelace
%T Generic Title
\end{verbatim}
\noindent
and vice versa.

@@

\subsection{Quotient Lenses}
Unfortunately, while our synthesis tool greatly simplifies the task of
programming bijective string lenses, not all bidirectional tranformations are bijective
in nature. For instance, users are typically uninterested in preserving
whitespace between words or between fields.  The order of author and title
fields is also likely irrelevant and there may be equivalent ways of writing
the same name:  ``Lovelace, Adan'' vs ``Ada Lovelace.''

When two instances of the same format differ only terms of such inessential
information, it is useful\bcp{Would it be worth spelling out the alternative
(i.e., what lenses would look like if they had to preserve all this stuff?)}
to treat them as identical and to map them into canonical representations.
Continuing our example, the following two \bibtex{} citations represent the same
logical object and as a result, should be translated to the same EndNote record.
\begin{verbatim}
@Book {Lovelace,
Author = "Ada Lovelace",
Title = {Generic Title},
}

@Book{     Lovelace,
Title = {Generic
Title},
Author = "Lovelace, Ada",     }
\end{verbatim}

@@

Sometimes there are multiple disjoint representations of the same data.
In such situations, the \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in the first format to the second.
For instance, assume that
the \codefont{comma\_name} format describes ``Lovelace, Ada''
and that the \codefont{space\_name} format describes ``Ada Lovelace''
and \codefont{c\_to\_s} is a function from the first to the second.  In this
case, the following instance of \kw{squash} creates the desired canonizer.

@@

One way to define the \codefont{c\_to\_s} canonization function used by squash
is simply to write it from scratch in some ordinary programming language.
However, our system will also synthesize such functions automatically from pairs
of regular languages and examples---here, \codefont{c\_to\_s} is the
\codefont{get} direction of a lens that can be synthesized by:
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Lovelace, Ada", "Ada Lovelace")}
let c_to_s = get l
\end{lstlisting}

@@

(* synthesize a lens that maps comma representation to space representation *)
let l = synth comma_name $\Leftrightarrow$ space_name using {("Lovelace, Ada", "Ada Lovelace")}
let c_to_s = get l

@@

Having identified QREs as a natural way of specifing equivalence relations on
regular expressions, a natural next step in defining quotient lenses is to map
between two QREs via a bijection between their; this is our
approach in defining {\em QRE lenses}.

More concretely, in order to define a language of QRE lenses, our approach is
to (1) define a language of bijections (Section \ref{subsec:bijective-lenses}),
and then (2) add quotients by allowing canonizers to be prepended or postpended
to bijections and (3) allowing composition of such quotient lenses via the
regular operators and functional composition
(Sections \ref{subsec:qre-lenses-syntax} and \ref{subsec:qre-lenses-semantics}).

However, we also have a secondary objective, which is to support synthesis of
lenses: When provided with QREs describing the source and target languages, we
would like to be able to generate quotient lenses automatically. One way to do
that is to generate canonizers $\canonizer(Q_1)$ and $\canonizer(Q_2)$ from QREs
$Q_1$ and $Q_2$ and then to synthesize bijective lenses between the kernel
languages for $Q_1$ and $Q_2$.

Unfortunately, the composition of two such quotient lenses does not have have
the form of a bijective lens with canonizers at the edges. Hence, an important
technical question is whether we give up expressiveness if we restrict
ourselves to this form. Thankfully, this turns out not to be an issue if
the bijections that we use to define QRE lenses are derived using a particular
set of combinators that we describe in the next section.

\subsection{Bijective String Lenses}
\label{subsec:bijective-lenses}
We define the set of \textit{bijective string lenses} to be the set of
bijections between regular languages constructed using the following combinators.

$$\ell := \mathit{id} \; (R) \sep \const(s_1, s_2) \sep  \swap(\ell_1, \ell_2)
\sep \ell_1 \cdot \ell_2 \; |  \; (\ell_1 \sep \ell_2) \sep \ell^* \; | \;
\ell_1 \; ; \;  \ell_2,$$ where $R$ ranges over regular expressions and $s$
ranges over character strings.

The $\mathit{\const}(s, t)$ lens replaces the string $s$ with $t$ in the source
data in the forward direction, and $t$ with $s$ in the target data in the
backward direction, while $\mathit{id}(R)$ applies the identity function to the
source and target  in $\mathcal{L}(R)$ in both directions. The composition lens
$\ell_1 \; ; \; \ell_2$ applies $\ell_1$ followed by then $\ell_2$ to the
source in the forward direction, and applies $\ell_2$ followed by $\ell_1$ to
the target in the backward direction. The lens $\ell_1 \cdot \ell_2$ first
splits the string $s$ into $s_1$ and $s_2$, applies $\ell_1$ and $\ell_2$ to
$s_1$ and $s_2$ to get $t_1$ and $t_2$ respectively, then concatenates $t_1$
and $t_2$ and returns $t_1 \cdot t_2$ as the final result in the forward
direction. $\ell_1 \cdot \ell_2$ operates similarly in the backward direction,
but with $s, s_1$ and $s_2$ substituted for $t, t_1$ and $t_2$. The
$\mathit{\swap} \; (\ell_1, \ell_2)$ lens operates like $\ell_1 \cdot \ell_2$,
except that it swaps $t_1$ and $t_2$ before concatenating the two for a final
result of $t_2 \cdot t_1$ in the forward direction. In the backward direction,
$\mathit{\swap}(\ell_1, \ell_2)$ first undoes the \swap, then behaves like
$cdot$. The $\ell_1 \; | \; \ell_2$ lens chooses to apply $\ell_1$ or $\ell_2$
depending on whether the source (resp. target) data is matched by $\ell_1$ or
$\ell_2$ in the forward (resp. backward direction). The $\ell^*$ lens splits the
string $s$ into strings $s_1, \ldots, s_n$, applies $\ell_1$ to each $s_i$ to
get $t_i$, and then concatenates each of the $t_i$'s for a final result of $t_1
\cdot \ldots \cdot t_n$ in the forward direction. The iteration lens $\ell^*$
operates similarly in the backward direction, but with $s, s_i$ substituted for
$t, t_i$.

The denotation of a lens $\ell_1$ is $\llbracket \ell_1 \rrbracket \subseteq
\mathit{String} \times \mathit{String}$. If $(s_1, s_2) \in \llbracket \ell_1
\rrbracket$, then $\ell_1$ maps between $s_1$ and $s_2$.
\begin{align*}
\llbracket R \rrbracket &= \{(r, r) \sep r \in \mathcal{L}(R)\}\\
\llbracket \const(r, s) \rrbracket &= \{(r, s)\}\\
\llbracket \swap(\ell_1, \ell_2) \rrbracket &= \{(s \cdot t, t' \cdot s') \sep
(s, s') \in \llbracket \ell_1 \rrbracket \text{ and } (t, t') \in \llbracket
\ell_2 \rrbracket\}\\
\llbracket \ell_1 \cdot \ell_2 \rrbracket &= \{(s \cdot t, s' \cdot t) \sep
(s, s') \in \llbracket \ell_1 \rrbracket \text{ and } (t, t') \in \llbracket
\ell_2 \rrbracket\}\\
\llbracket \ell_1 \sep \ell_2 \rrbracket &= \{(s \cdot t) \sep
(s, t) \in \llbracket \ell_1 \rrbracket \text{ or } (s, t) \in \llbracket
\ell_2 \rrbracket\}\\
\llbracket \ell^* \rrbracket &= \{(s_1 \cdot \ldots \cdot s_n, t_1 \cdot \ldots
\cdot t_n) \sep (s_i, t_i) \in \llbracket \ell \rrbracket \text{ for } 1
\leq i \leq n\}\\
\llbracket \ell_1 \; ; \; \ell_2 \rrbracket &= \{(r, t) \sep \text{there exists }s
\text{ such that } (r, s) \in \llbracket \ell_1 \rrbracket \text{ and } (s, t)
\in \llbracket \ell_2 \rrbracket\}
\end{align*}

Each bijective string lens $\ell$ has a type $\ell : R \Leftrightarrow S$ where
$R$ and $S$ are regular expressions. If $\ell : R \Leftrightarrow S$, then the
source language of $\ell$ is $\mathcal{L}(R)$ and the target language of $\ell$
is $\mathcal{L}(S)$. IThe final rule in the figure deserves special attention.
It states that a lens $\ell : R \Leftrightarrow S$ can also be considered to be
of type $\ell : R' \Leftrightarrow S'$ provided that $R$ (resp. $S$) can be
proven to be equivalent to $R'$ (resp. $S$) from the star-semiring axioms
(associativity and commutativity of | and $\cdot$, identity of the empty string
$\epsilon$, distibutivity of $\cdot$ over |, annihilative property of
$\varnothing$ with respect to $\cdot$, and that $R^* = \epsilon \; | \; R \cdot
R* = \epsilon \; | \; R^* \cdot R$ for all $R$):

@@

\noindent
This rule makes it significantly more difficult to synthesize a bijective lens
from its type.\sam{Should we explain why?}

@@

\subsection{Syntax of QRE Lenses}
\label{subsec:qre-lenses-syntax}
Having given a brief overview of the class of bijective string lenses, we now
introduce the class of QRE lenses. The language of QRE lenses is given by following
grammar:
$$ q := \kw {id}(R) \sep  \kw{const}(r, s) \sep q_1 \cdot q_2 \sep
\kw{swap}(q_1, q_2) \sep (q_1 \sep q_2) \sep q^* \sep q_1 \; ; \; q_2 \sep
\kw{lquot}(q, \ell) \sep \kw{rquot}(\ell, q)$$ where $R$ ranges over
regular expressions and $\ell$ ranges over bijective lenses.

@@

\subsection{Semantics of QRE Lenses}
\label{subsec:qre-lenses-semantics}
Each QRE lens $q$ has a type $q : Q_1 \Leftrightarrow Q_2$ where $Q_1, Q_2$ are
QREs. If $q : Q_1 \Leftrightarrow Q_2$, then the source format is described by
$W(Q_1)$ and the canonical set of representative for the source data is
described by $K(Q_1)$.
Similarly, the target format is described by $W(Q_2)$ and the canonical set of
representatives for the target data is described by $K(Q_2)$. The underlying
lens of $q$ is a bijective lens $\ell : K(Q_1) \Leftrightarrow K(Q_2)$.
